---
title: "EMOTIKON - Preprocessing demos: forcats package"
author: "Reinhold Kliegl"
date: "2022-07-18 (updated: `r Sys.time()`)"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: no
    code_folding: show
editor_options: 
  chunk_output_type: console
---

# Background

See McNamara & Horton (2017) in /literature. 

# Setup

+ Read a subset of EMOTIKON data
+ Compute zscores for each test

```{r}
library(lme4)
library(tidyverse)
library(magrittr)

dat <- 
  readRDS("data/EmotikonSubset.rds") |> 
  group_by(Test)|>
  mutate(zScore = as.numeric(scale(score)))

dat |>
  group_by(Test) |> 
  summarise(N=n(), M=mean(score), SD=sd(score), zM=mean(zScore), zSD=sd(zScore))

dat |> 
  ggplot(aes(x=age, y=zScore, group=Sex, color=Sex)) +
  facet_grid(. ~ Test) +
#  geom_point() +
  geom_smooth(method="lm") +
  scale_colour_manual("", values=c("red", "blue")) +
  theme_bw()


```

# Some basic about the critical factors

## fct_count

Tibble with levels and instances at each level. 

```{r}
fct_count(dat$Cohort)
n_Cohort <- fct_count(dat$Cohort)$n  # extract a variable

fct_count(dat$Sex)

fct_count(dat$Test)
```

## fct_unique

Vector with levels as values.

```{r}
fct_unique(dat$Cohort)
levels(dat$Cohort)
```

# Generate a default figure for study

We need 

+ table of means
+ figure

```{r}
# table of means
t1_AST <- 
  dat %>%
  mutate(AgeBins = cut( age, breaks = c(7.8, seq(8.2, 9.0, 0.25), 9.3)))|>
  group_by(AgeBins, Sex, Test)|>
  summarise(N=n(), Age=mean(age),  M=mean(zScore), SD=sd(zScore), 
            CI=1.96*SD/sqrt(N), .groups="drop") 
t1_AST

# figure
fig1 <- 
  dat %>%
  ggplot(aes(x=age, y=zScore, group=Sex, shape=Sex)) +  facet_grid(. ~ Test) +
  geom_smooth(method="lm", formula=y ~ poly(x,1), se=FALSE,  colour="black", size=.75) + 
  geom_point(data=t1_AST, aes(x=Age, y=M), fill="white") + 
  scale_shape_manual(values=c(16,21)) +
  scale_x_continuous("Age", breaks=seq(8,9,.5)) +
  scale_y_continuous("z-score [± SD]") + 
  geom_hline(aes(yintercept=0), color="black") + 
  coord_cartesian(xlim=c(8.0, 9.1), ylim=c(-0.75,+0.75)) +
  theme_bw() +  theme(legend.justification=c(0,1), legend.position=c(0, 1), 
  legend.background=element_rect(fill="white", color="black"), legend.title=element_blank())
fig1
```

What is **not** nice about this graph?

+ new labels for components
+ new labels for sex
+ different order for sex

# Recode

Change the **label** of factor levels with new factor; could also overwrite the old one

+ fct_recode()

```{r}
dat <-  
  dat |> 
  mutate(Component = 
         fct_recode(Test, Endurance = "Run", 
                          Coordination = "Star_r", 
                          Speed = "S20_r",
                          PowerLOW = "SLJ", 
                          PowerUP = "BPT"),
         Sex = fct_recode(Sex, Girls = "female", Boys = "male"))
dat
```



```{r}
# table of means
t2_AST <- 
  dat %>%
  mutate( AgeBins = cut( age, breaks = c(7.8, seq(8.2, 9.2, 0.25), 9.3)))|>
  group_by(AgeBins, Sex, Component)|>
  summarise(N=n(), Age=mean(age),  M=mean(zScore), SD=sd(zScore), 
            CI=1.96*SD/sqrt(N), .groups="drop") 
t2_AST

# figure
fig2 <- 
  dat %>%
  ggplot(aes(x=age, y=zScore, group=Sex, shape=Sex)) +  facet_grid(. ~ Component) +
  geom_smooth(method="lm", formula=y ~ poly(x,1), se=FALSE,  colour="black", size=.75) + 
  geom_point(data=t2_AST, aes(x=Age, y=M), fill="white") + 
  scale_shape_manual(values=c(16,21)) +
  scale_x_continuous("Age", breaks=seq(8,9,.5)) +
  scale_y_continuous("z-score [± SD]") + 
  geom_hline(aes(yintercept=0)) + 
  coord_cartesian(xlim=c(8.0, 9.1), ylim=c(-0.75,+0.75)) +
  theme_bw() +  theme(legend.justification=c(0,1), legend.position=c(0, 1), 
  legend.background=element_rect(fill="white", color="black"), legend.title=element_blank())
fig2
```

# Reverse or relevel

Reverse the order of factor levels

+ fct_rev()       # reverse
+ fct_relevel()   # relevel
+ fct_infreq()    # frequency
+ fct_inorder()   # in order of appearance

## fct_rev()

```{r}
dat <- dat |>  mutate(Sex = fct_rev(Sex))

# table of means
t3_AST <- 
  dat %>%
  mutate( AgeBins = cut( age, breaks = c(7.8, seq(8.2, 9.2, 0.25), 9.3)))|>
  group_by(AgeBins, Sex, Component)|>
  summarise(N=n(), Age=mean(age),  M=mean(zScore), SD=sd(zScore), 
            CI=1.96*SD/sqrt(N), .groups="drop") 
t3_AST

# figure
fig3 <- 
  dat %>%
  ggplot(aes(x=age, y=zScore, group=Sex, shape=Sex)) +  facet_grid(. ~ Component) +
  geom_smooth(method="lm", formula=y ~ poly(x,1), se=FALSE,  colour="black", size=.75) + 
  geom_point(data=t3_AST, aes(x=Age, y=M), fill="white") + 
  scale_shape_manual(values=c(16,21)) +
  scale_x_continuous("Age", breaks=seq(8,9,.5)) +
  scale_y_continuous("z-score [± SD]") + 
  geom_hline(aes(yintercept=0)) + 
  coord_cartesian(xlim=c(8.0, 9.1), ylim=c(-0.75,+0.75)) +
  theme_bw() +  theme(legend.justification=c(0,1), legend.position=c(0, 1), 
  legend.background=element_rect(fill="white", color="black"), legend.title=element_blank())
fig3
```

This figure corresponds to Figure 1 in paper. 


# fct_relevel() 

```{r}
levels(dat$Component)
dat <- dat |>  mutate(Compx = fct_rev(Component))
levels(dat$Compx)

dat <- dat |>  mutate(Comp2 = fct_relevel(Component, c("PowerUP","Endurance", "Coordination",  "Speed",  "PowerLOW" ) ))
levels(dat$Comp2)

t4_AST <- 
  dat %>%
  mutate( AgeBins = cut( age, breaks = c(7.8, seq(8.2, 9.2, 0.25), 9.3)))|>
  group_by(AgeBins, Sex, Comp2)|>
  summarise(N=n(), Age=mean(age),  M=mean(zScore), SD=sd(zScore), 
            CI=1.96*SD/sqrt(N), .groups="drop") 
t4_AST

# figure
fig4 <- 
  dat %>%
  ggplot(aes(x=age, y=zScore, group=Sex, shape=Sex)) +  facet_grid(. ~ Comp2) +
  geom_smooth(method="lm", formula=y ~ poly(x,1), se=FALSE,  colour="black", size=.75) + 
  geom_point(data=t4_AST, aes(x=Age, y=M), fill="white") + 
  scale_shape_manual(values=c(16,21)) +
  scale_x_continuous("Age", breaks=seq(8,9,.5)) +
  scale_y_continuous("z-score [± SD]") + 
  geom_hline(aes(yintercept=0)) + 
  coord_cartesian(xlim=c(8.0, 9.1), ylim=c(-0.75,+0.75)) +
  theme_bw() +  theme(legend.justification=c(0,1), legend.position=c(0, 1), 
  legend.background=element_rect(fill="white", color="black"), legend.title=element_blank())
fig4
```


# `fct_infreq()` and `fct_inorder()´ 

We can sort factor levels by their frequency or by their order of appearance in the data frame. 

```{r}
datx <- tibble(f = factor(c("b", "b", "a", "c", "c", "c")))
datx$f
fct_inorder(datx$f)
fct_infreq(datx$f)
fct_rev(fct_infreq(datx$f))

datx <- datx |> mutate(g = fct_inorder(f), h = fct_infreq(f))
datx$f
datx$g
datx$h
```

## `fct_collapse()`


```{r}
dat <- dat |> mutate(Comp4 = fct_collapse(Component, 
                                     'Endurance + PowerUP' = c("Endurance", "PowerUP"), 
                                     'Physical skill' = c("Coordination", "Speed", "PowerLOW"))) 
levels(dat$Comp4)


t6_AST <- 
  dat %>%
  mutate( AgeBins = cut( age, breaks = c(7.8, seq(8.2, 9.2, 0.25), 9.3)))|>
  group_by(AgeBins, Sex, Comp4)|>
  summarise(N=n(), Age=mean(age),  M=mean(zScore), SD=sd(zScore), 
            CI=1.96*SD/sqrt(N), .groups="drop") 
t6_AST

# figure
fig6 <- 
  dat %>%
  ggplot(aes(x=age, y=zScore, group=Sex, shape=Sex)) +  facet_grid(. ~ Comp4) +
  geom_smooth(method="lm", formula=y ~ poly(x,1), se=FALSE,  colour="black", size=.75) + 
  geom_point(data=t6_AST, aes(x=Age, y=M), fill="white") + 
  scale_shape_manual(values=c(16,21)) +
  scale_x_continuous("Age", breaks=seq(8,9,.25)) +
  scale_y_continuous("z-score [± SD]") + 
  geom_hline(aes(yintercept=0)) + 
  coord_cartesian(xlim=c(8.0, 9.1), ylim=c(-0.75,+0.75)) +
  theme_bw() +  theme(legend.justification=c(0,1), legend.position=c(0, 1), 
  legend.background=element_rect(fill="white", color="black"), legend.title=element_blank())
fig6
```


# A bit more sophisticated ...

## `fct_reorder()`

```{r  }
ix <- sample(1:10000, 5000)
dat2 <- dat[ix,]

# We could also fct_reorder -- by default small first
dat2$Comp3 <- fct_reorder(dat2$Component, dat2$Component, length)
fct_count(dat2$Comp3)  # does work!

# also other statistics, e.g., median
dat|>group_by(Component)|>summarise(N=n(), median=median(zScore, na.rm=TRUE))
dat|>group_by(Component)|>summarise(N=n(), mean=mean(zScore, na.rm=TRUE))

t5_AST <- 
  dat2 %>%
  mutate( AgeBins = cut( age, breaks = c(7.8, seq(8.2, 9.2, 0.25), 9.3)))|>
  group_by(AgeBins, Sex, Comp3)|>
  summarise(N=n(), Age=mean(age),  M=mean(zScore), SD=sd(zScore), 
            CI=1.96*SD/sqrt(N), .groups="drop") 
t5_AST

# figure
fig5 <- 
  dat %>%
  ggplot(aes(x=age, y=zScore, group=Sex, shape=Sex)) +  facet_grid(. ~ Comp3) +
  geom_smooth(method="lm", formula=y ~ poly(x,1), se=FALSE,  colour="black", size=.75) + 
  geom_point(data=t5_AST, aes(x=Age, y=M), fill="white") + 
  scale_shape_manual(values=c(16,21)) +
  scale_x_continuous("Age", breaks=seq(8,9,.5)) +
  scale_y_continuous("z-score [± SD]") + 
  geom_hline(aes(yintercept=0)) + 
  coord_cartesian(xlim=c(8.0, 9.1), ylim=c(-0.75,+0.75)) +
  theme_bw() +  theme(legend.justification=c(0,1), legend.position=c(0, 1), 
  legend.background=element_rect(fill="white", color="black"), legend.title=element_blank())
fig5
```

## fct_reorder2

+ `fct_reorder()` reorders levels by their relationship (length, mean, median) with another variable (see above).
+ `fct_reorder2()` reorders levels by their final values when plotted with two other variables.

This corresponds (roughly) to Figure 4: Cohort x Component in paper, but tests appear in order of mean `zScore` of last cohort (2019). The example illustrates how this can be done in the `ggplot()` command.

The initial result was almost, but not exactly what expected:  Endurance was not lowest, but in third position because of some surprising prediction of the cubic prediction. To get the order using the 2019 means, `geom_point()` must appear before `geom_smooth()`. 

Of course, we could have reordered already when constructing the table.  

```{r fig.width=7, fig.height=3.5}
t_CT <- dat|>
  mutate(cohort = as.numeric(as.character(Cohort)))|>
  group_by(cohort, Component, Child)|>  
  summarise(N=n(), M=mean(zScore))|>
  group_by(cohort, Component)|>
  summarise(N=n(), zScore=mean(M), SD=sd(M), CI=1.96*SD/sqrt(N), .groups="drop")|>
  mutate(Comp6 = fct_reorder2(Component, cohort, zScore))

tail(t_CT)

# n of children per cohort
nChild <-
  dat|>
  group_by(Cohort)|>
  summarise(N=length(unique(Child)))

fig8 <-
  dat %>%
  mutate(cohort = as.numeric(as.character(Cohort)))|>
  ggplot(aes(x=cohort, y=zScore, 
             group=Component, color=Component)) + 
  geom_point(data=t_CT, aes(x=cohort, y=zScore, group=Comp6, color=Comp6), fill="white") + 
  geom_smooth(method="lm", formula=y~poly(x,3), se=FALSE) +
  scale_shape_manual(values=c(16,21)) +
  scale_x_continuous("Cohort", breaks=seq(2011, 2019)) +
  scale_y_continuous("z-score [± SD]") + 
#  geom_errorbar(data=t_CT, aes(ymax=zScore+CI, ymin=zScore-CI), width=0.05) +
  scale_colour_manual("Component", values=c("red", "orange", "green", "blue", "violet")) +
  geom_hline(aes(yintercept=0)) + 
  coord_cartesian(ylim=c(-0.2,+0.2)) +
  theme_bw(base_size=14) +
  theme(legend.position="right") +
  annotate("text", x=2011+.30, y= -.18, label="N of children:", size=3) +
  annotate("text", x=2011:2019, y= -0.195, label=nChild$N, size=3)
fig8
```

The results show the expected loss for `Endurance` and gain in `Speed`. Of interest are also the curvilinear trends for the other three components, especially `PowerUP`. The differences between components show up as reliable cohort-related VCs. The size of the VC corresponds with the size of the effects shown in the figure. Note the difference in effect size is much smaller than the effect size associated with the age of children within the third grade.

## Reorder in a facet plot conditional on the values in a specific facet

```{r}
fm1 <- lmer(Reaction ~ Days + (Days|Subject), sleepstudy)
rr1 <- ranef(fm1)
dd <- as.data.frame(rr1)

# Default: Subj ordered by "(Intercept)"
ggplot(dd, aes(y=grp,x=condval)) +
        geom_point() + facet_wrap(~term,scales="free_x") +
        geom_errorbarh(aes(xmin=condval -2*condsd,
                           xmax=condval +2*condsd), height=0) +
        xlab("Conditional mean") + ylab("Subject") +
        theme_bw()
 
# Subj ordered by ...
ord_subj <- dd |> filter(term=="Days") |> arrange(condval) |> pull(grp) 

ggplot(dd, aes(y=factor(grp, levels=ord_subj), x=condval)) +
      geom_point() + facet_wrap(~term,scales="free_x") +
      geom_errorbarh(aes(xmin=condval -2*condsd,
                         xmax=condval +2*condsd), height=0) +
      xlab("Conditional mean") + ylab("Subject") +
      theme_bw()
```


# Appendix

```{r}
sessionInfo()
```

