---
title: "DSSS-02: Reading Participants' Data"
author: "Reinhold Kliegl"
date-format: iso
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    number-sections: true
    fig-width: 10
    fig-height: 7
    fig-format: svg
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
library(tidyverse)
library(readxl)   # this package allows us to read xlsx files
library(lubridate)


library(summarytools) # dfSummary(), descr(), freq()
```

# RK

## Context
Simulation of simple reaction times from a picture-naming experiment. Fifty subjects name  75 different pictures (items) under four transparency conditions (i.e., 300 trials per subject; total N = 15,000). The greyness of the screen background is randomly sampled from a normal distribution for each trial. 

## Issues
1. Theory-guided contrast specification for four-level factor
2. Specification of crossed-random factor linear mixed models
3. Visualization of partial effects of covariate x factor interactions

## Variables in file
    Subj : subject id; random factor
    Item : item id; random factor
    Trnsp: transparency of item (levels: A = 100%, B = 75%, C = 50 %, D = 25%); 
           within-subject and within-item fixed factor
    scrb : screen background; within-subject and within-item covariate
    rt   : reaction time, dependent variable

## Preprocessing

### Read the data

```{r}
dat <- read_csv("data/simul_data.csv")
```

### A few quick checks

```{r}
view(dfSummary(dat))

descr(dat, stats="common")  
descr(dat, stats="fivenum") 
```

### Fix `Subj` and `Item` type and levels

+ same length of levels by left padding with zeros
+ convert to factor

```{r}
dat <- 
  dat |> 
   mutate(
    Subj = as_factor(paste0("S", str_pad(Subj, width = 2, side = "left", pad = "0"))),
    Item = as_factor(paste0("I", str_pad(Item, width = 2, side = "left", pad = "0")))
    )

view(dfSummary(dat))
```

# N.N.


# Appendix

```{r}
sessionInfo()
```

