---
title: "DSSS-02: Reading Students' Data"
author: "Reinhold Kliegl"
date: "2022-02-17 (updated: `r Sys.Date()`)"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: no
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
library(tidyverse)
library(readxl)
library(lubridate)


library(summarytools)

```

# Exploring students' data

## Danila

### Read the data

```{r}
datw_dk <- read_csv("data/Kramarenko_1_TSLA.csv")
datw_dk
```

### A few quick checks

```{r}
view(dfSummary(datw_dk))

descr(datw_dk)  
```

We do not get a statistic for `Date` because the variable is not of type `[numeric]`

### Graph

Quick and dirty ...

```{r}
dat_dk <- datw_dk |> pivot_longer(2:6, names_to="Course", values_to="Price")
dat_dk

dat_dk |> 
  filter(Course %in% c("Low", "High")) |> 
  ggplot(aes(x=Date, y=Price, group=Course, color=Course)) +
  geom_point() + geom_line() +
  scale_x_date("", date_breaks = "1 year", date_minor_breaks= "3 months", date_labels= "%Y-%m") +
  scale_y_continuous("Price") +
  scale_color_manual("Course", values=c("blue", "red")) +
  theme_bw()
```

We can do better with a few changes to the above command.

Alternatives
  + transfoming
  + smoothing
  + ...
    

## Victoria

Participants with media experience rating credibility of true and fake news about hot topics (abortion, LGBQ, death penalty). Subject number was added as shown in class.


```{r}
datw_vv <- read_xlsx("data/Vziatysheva_1_MEXP.xlsx")
datw_vv
```

### A few quick checks

```{r}
view(dfSummary(datw_vv))

descr(datw_vv)  
freq(datw_vv)
```

### Graph

```{r}

```


## Hamid Keshavarz

### Read the data

This is a challenge for me, too. Happy to learn how we deal with this format.

```{r}
datw_001_hk <- read.csv("data/Keshavarz_001_Mayer.csv")
datw_001_hk

# Hm?

datw_002_hk <- read.csv("data/Keshavarz_001_Mayer.csv")
datw_002_hk
```


### A few quick checks


```{r}
view(dfSummary(datw_hk))

descr(datw_hk)  
freq(datw_hk)
```


# EMOTIKON - physical fitness 

An example of a relational data base. 

+ Child
+ School
+ Community

```{r}



```

# Appendix

```{r}
sessionInfo()
```

