---
title: "DSSS-02: Reading Students' Data"
author: "Reinhold Kliegl"
date: "2022-02-17 (updated: `r Sys.Date()`)"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: no
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
library(tidyverse)
library(readxl)   # this package allows us to read xlsx files
library(lubridate)


library(summarytools) # dfSummary(), descr(), freq()
```

# Exploring students' data

## Danila Kramarenko

### Read the data

```{r}
datw_dk <- read_csv("data/Kramarenko_1_TSLA.csv")
datw_dk
```

### A few quick checks

```{r}
view(dfSummary(datw_dk))

descr(datw_dk, stats="common")  
descr(datw_dk, stats="fivenum") 
```

We do not get a statistic for `Date` because the variable is not of type `[numeric]`

### Graph

For an efficient generation of plots we usually need the data in wide format.

```{r}
dat_dk <- datw_dk |> pivot_longer(2:5, names_to="StockStatistic", values_to="Price")
dat_dk

dat_dk |> 
  filter(StockStatistic %in% c("Low", "High") & Date < "2013-01-01") |> 
  ggplot(aes(x=Date, y=Price, group=StockStatistic, color=StockStatistic)) +
  geom_smooth() +
#  geom_point() + geom_line() +
  scale_x_date("", date_breaks = "1 year", date_minor_breaks= "3 months", date_labels= "%Y-%m") +
  scale_y_log10("Price") +
  scale_color_manual("Stock statistic", values=c(  "green", "red")) +
  theme_bw(base_size=14)
```

We can do better with a few changes to the above command.

Alternatives
  + transfoming
  + smoothing
  + ...
  
  
For correlations we need to variables to be correlated in the wide format. For example,...  
    
```{r}
cor(datw_dk[, 2:5])
```

The high correlations are spuriously high due to the change in stocks over years. We could try to remove the overall trend and check the correlations then.

## Victoria Vziatysheva

Participants with media experience rating credibility of true and fake news about hot topics (abortion, LGBTQ, death penalty). Subject number was added as shown in class.


```{r}
datw_vv <- read_xlsx("data/Vziatysheva_1_MEXP.xlsx")
datw_vv
```

### A few quick checks

```{r}
view(dfSummary(datw_vv))

descr(datw_vv, stats="common")  
freq(datw_vv)
```

### Graph

```{r}

```


## Hamid Keshavarz

### Read the data

This is a challenge for me, too. Happy to learn how we deal with this format.

```{r}
datw_1_hk <- read.csv("data/Keshavarz_1_Mayer.csv")
datw_1_hk

# Hm?

datw_2_hk <- read.csv("data/Keshavarz_1_Mayer.csv")
datw_2_hk
```

### A few quick checks

```{r}
view(dfSummary(datw_1_hk))

freq(datw_1_hk)
```

## Bella Mamedova

```{r}
datw_bm <- read_xlsx("data/Mamedova_1_GermanCredit.xlsx") 
datw_bm
```

### A few quick checks

```{r}
view(dfSummary(datw_bm))

descr(datw_bm)  
freq(datw_bm)
```

### Graph

```{r}

```

# Anastasia Yanechko

```{r}
datw_ay <- read.csv("data/Yanechko_1_Anime_v2.csv")
datw_ay
```

### A few quick checks

```{r}
view(dfSummary(datw_ay))

descr(datw_ay)  
freq(datw_ay)
```

### Graph

```{r}

```

# EMOTIKON - physical fitness 

An example of a relational data base. 

+ Child
+ School
+ Community

```{r}



```

# Appendix

```{r}
sessionInfo()
```

