---
title: "DSSS-03: Tidyverse Commands"
author: "Reinhold Kliegl"
date: "2022-07-18 (updated: `r Sys.Date()`)"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: no
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

# Setup

```{r}
library(tidyverse)
library(readxl)   # this package allows us to read xlsx files
library(lubridate)

library(summarytools) # dfSummary(), descr(), freq()
```

# Read data 

We use Reinhold Kliegl's data.

```{r}
dat <- read.csv("participants/Kliegl.csv")
```

# Preprocessing

```{r}

```


# tidyverse
 
Vocabulary -- is about 10 commands

## Commands working on columns (variables)

+ `select()` -- extract a subset of columns
+ `relocate()` -- adjust position of variable

## Commands works on rows (usually observations)

+ `filter()` -- extract a subset of rows 
+ `arrange()` -- orders the observations in the data frame
+ `mutate()` -- add new variables/columns, unless you overwrite an existing one.
    
## Logical operators

+ ==, >, >, >=, >=
+ |, &,  %in%
+ is.na()

    
## Arithmetic operators

+ +, -, *, /, ^
+ log(), log(2), log10()
+ lead(), lag(), lag2()
+ cumsum()
    

```{r}

```


# Appendix

```{r}
sessionInfo()
```

