---
title: "DSSS-01: Introduction"
author: "Reinhold Kliegl"
date: "2022-07-18 (updated: `r Sys.Date()`)"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: no
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

# Background

## Theory

< Describe the research questions and data>

### Proposition 1


### Proposition 2


## Variables

< Dictionary of variables >

# Setup

R needs packages. You can install the packages needed for this script by running the `install.packages()` command in the next R chunk once. This command retrieves the packages from the web and installs them in a library on your computer. After you installed the packages, you can actually delete this chunk. The `library()`  command fetches the packages from the library on your computer and makes the commands in the packages available for the currect session. 


```{r eval=FALSE}
install.packages(c("tidyverse", "readxl", "WriteXLS", "summarytools" ))
```

The following commands retrieve the packages from the library on your computer. This needs to be done at the beginning of every script.

```{r}
library(tidyverse)
library(readxl)
library(WriteXLS)
library(summarytools)
```

# Toy example data

+ Generate data
+ Write data to Excel spreadsheet

```{r}
set.seed(42)
x <- round(   runif(300, min=0, max=10  )   )
length(x)

X <- matrix(x, nrow=20, byrow=FALSE)

df0 <- data.frame(Subj=1:20)

df <- data.frame(Subj=1:20, X)

WriteXLS(df, "./data/DSSS_01.xlsx")
```

# Read data

Now we can read the data we generated back into R. 

```{r}
df2 <- read_xlsx("./data/DSSS_01.xlsx")
```

# Preprocessing

```{r}
df3 <- 
  df2 %>% 
  mutate(Subj=paste0("S", Subj))
df3

df3 |> ggplot(aes(x=Subj, y=X1, group=1)) + geom_line() 
# Ugly graphic!


df4 <- 
  df2 |> 
  mutate(Subj = paste0("User ", str_pad(Subj, width=2, side="left", pad="0")))

df4
df4 |> 
  ggplot(aes(x=Subj, y=X1, group=1)) + 
  geom_line() + 
  xlab("")
# Much better graphic!

# Alternative, not-tidyverse style of generating the same graph
ggplot(df4, aes(x=Subj, y=X1, group=1)) + geom_line() + xlab("")
```

# Descriptive statistics

```{r}
descr(df2)
```


# Graphs

```{r}
df4 |> ggplot(aes(x=X1, y=X2)) + geom_point() + theme_bw()
```


# Inferential statistics

For example, correlations between first five variables. 

```{r}
df4 |>  select(X1:X5) |> cor()
```

# Appendix

# How to we change the default RMarkdown script?

We can google "changing default RMarkdown script". This page is offered:

https://community.rstudio.com/t/changing-default-rmd-script/36426 

Not very helpful, unfortunately. At least, I cannot see the file with the default code.

# What packages did we use in this session?

```{r}
sessionInfo()
```

