---
title: "Data Science for Social Science - 01"
author: "Reinhold Kliegl"
date: "2022-02-10 (updated: `r Sys.Date()`)"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: no
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

# Background

## Theory
< Describe the research questions and data>

### Proposition 1


### Proposition 2


## Variables

< Dictionary of variables >dasflasjemfrklwqejrtoqwjklöarjäoakwerlÖQWE

# Setup

R needs packages. You can install the packages needed for this script by running the `install.packages()` command in the next R chunk once. This command retrieves the packages from the web and installs them in a library on your computer. After you installed the packages, you can actually delete this chunk. The `library()`  command fetches the packages from the library on your computer and makes the commands in the packages available for the currect session. 


```{r eval=FALSE}
install.packages(c("tidyverse", "readxl", "WriteXLS", "summarytools" ))
```

The following commands retrieve the packages from the library on your computer. This needs to be done at the beginning of every script.

```{r}
library(tidyverse)
library(readxl)
library(WriteXLS)
library(summarytools)
```

# Generate data

+ Generate data
+ Write to Excel spreadsheet

```{r}
x <- round(runif(300, min=0, max=10))
X <- matrix(x, nrow=20)
df <- data.frame(Subj=1:20, X)

WriteXLS(df, "./data/DSSS_01.xlsx")
```

# Read data

```{r}
df2 <- read_xlsx("./data/DSSS_01.xlsx")
```

# Transformations

```{r}
df2 <- 
  df2 |> 
  mutate(Subj = paste0("S", str_pad(Subj, width=2, side="left", pad="0")))

df2
```

# Descriptive statistics

```{r}
descr(df2)
```


# Graphs

```{r}
df2 |> ggplot(aes(x=X1, y=X2)) + 
  geom_point()
```


# Inferential statistics

For example, correlations between first five variables. 

```{r}
df2 |>  
  select(X1:X5) |> 
  cor()
```

# Appendix

```{r}
sessionInfo()
```

