---
title: "Janis Fiedler: Collaborative Goals for PA and FVI (Family)"
author: "Reinhold Kliegl"
date: "2022-09-17 (last revised: `r format(Sys.time())`)"
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    number-sections: true
    fig-width: 8
    fig-height: 6
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(arrow)
library(tidyverse)
library(easystats)
library(summarytools)
library(lme4)
```

# Background

The aim of this paper is to evaluate the association of self-selected collaborative goal achievement with PA and FVI changes from pre-post in families.
Additionally, the contribution of different family members to collaborative goal achievement will be shown descriptively.

File `df_fam` includes data of the 3 week mobile health intervention period during the SMARTFAMILY trial. Families set self selected goals and received timely feedback on the SMARTFAMILY app.
Pre and Post Data - which consists of a one week measurement each - is still missing from the excel file but I will provide it until the workshop.

File `df_ind` includes the same data only not for the whole family but for each family member. Here, I will add sex and age of the participants to the data.

The Data has not been screened yet for unrealistic values (e.g. if participants tyxped 50 instead of 5 portions into the application) but the raw data is available and will be screened for mistakes until the workshop.


# Preprocessing

## RK style guide

+ Random factors should be factors, not integers!
+ Factors capitalized, continuous variables in lower case
+ `Subj` instead of `ID`; prefix subject and item numbers with `S` and `I` 
+ Short variable names and factor levels, but respect tradeoff with mmemonic value
+ Labels should be of equal length
+ Hint: Leave out vowels in names and labels
+ See Jenny Bryan's [Naming Things](https://docplayer.net/55248970-Naming-things-prepared-by-jenny-bryan-for-reproducible-science-workshop.html)

## Packages

```{r}
library(easystats)
library(tidyverse)
library(lme4)
library(hypr)
library(summarytools)

# respecting color vision deficiency
cbPalette <- c( "#0072B2", "#D55E00", "#009E73", "#CC79A7",
                "#F0E442", "#56B4E9", "#999999", "#E69F00")
```

## Read and transform data

+ Assumption: each row is one person
+ Bew43T3: excluded MoMo_T3 In meiner Wohnumgebung fahren Autos
+ atp: active transport; mean of 11 items (Bew37T3 to Bew48T3; each coded 1 to 4)
+ wuSp: sum of 5 binary items (WUsport1_T3 to WUsport5_T3; 1="ja", 2="nein"; 10-sum = n of "ja")
+ SES: assign missing to medium

```{r}
library(labelled)

datBW <- 
    readRDS("KLOS.rds") |> 
    remove_labels() |> 
    select(starts_with("Bew"), -Bew43T3) |> 
    rowid_to_column(var="Subj") |> 
    mutate(Subj = factor(paste0("S", str_pad(Subj, width = 4, side = "left", pad = "0")))) |> 
    rowwise() |> 
    mutate(atp = mean(c_across(starts_with("Bew")), na.rm=TRUE)) 

datWU <- 
  readRDS("KLOS.rds") |> 
  remove_labels() |> 
  select(starts_with("WUsport")) |> 
  rowid_to_column(var="Subj") |> 
  mutate(Subj = factor(paste0("S", str_pad(Subj, width = 4, side = "left", pad = "0")))) |> 
  rowwise() |> 
  mutate( wuSp = 10-sum(c_across(starts_with("WUsport")))) 

dat_w <- 
  readRDS("KLOS.rds") |> 
  remove_labels() |> 
  select(-Bew43T3) |> 
  rowid_to_column(var="Subj") |> 
  mutate(Subj = factor(paste0("S", str_pad(Subj, width = 4, side = "left", pad = "0"))),
         SDEses_T3 = ifelse(is.na(SDEses_T3), 2, SDEses_T3),
         SES = factor(SDEses_T3, labels=c("low", "medium", "high")),
         ses = ifelse(is.na(SDEses_score_T3), 14, SDEses_score_T3),
         ses_c = ses - 14,
         wuOrt = ifelse(is.na(WUort1_jj_T3), 2004, WUort1_jj_T3),
         wuOrt_c = wuOrt-2004) |> 
  rowwise() |> 
  mutate(atp = mean(c_across(starts_with("Bew")), na.rm=TRUE),
         atp = ifelse(is.na(atp), 2.9, atp),
         atp_c = atp - 2.5,
         wuSp =10 - sum(c_across(starts_with("WUsport")))) |> 
  select(Subj,  SES, ses, ses_c, wuSp, wuOrt, wuOrt_c,  atp, atp_c,
         fuss_3=iafussT3, rad_3=iaradT3)

#stview(dfSummary(dat_w))

dat <- 
  dat_w |> 
  pivot_longer(10:11, names_to=c("Type", "Time"), names_sep="_", values_to="rating") |> 
  mutate(Type=fct_recode(Type, "foot"='fuss', "bicycle"='rad')) |> 
  filter(!is.na(rating))

MASS::boxcox(rating ~ atp+Type, data=dat) # looks good

#stview(dfSummary(dat))
```

**Issues**

+ Missing values for `wuSp`
+ Replacement of missing values for other covariates
+ Only two dependent measures so far, integrate T4!

# Descriptives (smooths)

## SES

```{r}
dat |> 
  ggplot(aes(x=ses, y=rating, group=Type, color=Type)) +
#  geom_smooth() +
  geom_smooth(method="lm") +
  scale_colour_manual("Type", values=cbPalette) +
  scale_x_continuous("Socioeconomic status") +
  scale_y_continuous("Rating") +
  theme_bw()  +
  coord_cartesian(xlim=c(1, 25), ylim=c(1, 6))
```

## Move to location

```{r}
dat |> 
  ggplot(aes(x=wuOrt, y=rating, group=Type, color=Type)) +
#  geom_smooth() +
  geom_smooth(method="lm") +
  scale_colour_manual("Type", values=cbPalette) +
  scale_x_continuous("Move to location") +
  scale_y_continuous("Rating") +
  theme_bw()  +
  coord_cartesian(xlim=c(1985, 2017), ylim=c(1, 6))
```

## Active transport

```{r}
dat |> 
  ggplot(aes(x=atp, y=rating, group=Type, color=Type)) +
  geom_smooth() +
#  geom_smooth(method="lm") +
  scale_colour_manual("Type", values=cbPalette) +
  scale_x_continuous("Active transport") +
  scale_y_continuous("Rating") +
  theme_bw()  +
  coord_cartesian(xlim=c(1, 4), ylim=c(1, 6))
```

# Linear mixed models

## Contrasts and trends

+ `contr.sum` uses ‘sum to zero contrasts’, that is returns the difference of factor level means to Grand Mean (except for the last level); Grand Mean is mean of condition means. 
+ Not much evidence for more than linear trends


```{r}
contrasts(dat$Type) <- contr.sum(2)
contrasts(dat$Type)
```

## _A priori_ LMM

```{r}
m_voi1 <- lmer(rating ~ 1 + Type*(ses_c+wuOrt_c + atp_c) + (1 | Subj), data=dat, REML=FALSE,
              control=lmerControl(calc.derivs=FALSE))
print(summary(m_voi1), corr=FALSE)
```

## _Post-hoc_ LMM

```{r}
m_voi2 <- lmer(rating ~ 1 + Type/(ses_c+wuOrt_c + atp_c) + (1 | Subj), data=dat, REML=FALSE,
              control=lmerControl(calc.derivs=FALSE))
anova(m_voi1, m_voi2)

print(summary(m_voi2), corr=FALSE)
```


# Appendix

```{r}
sessionInfo()
```

