---
title: "Introduction to R and RStudio: Correlations"
author: "Reinhold Kliegl"
date: "2022-07-18 (last revised: `r format(Sys.time())`)"
output: 
    html_document:
        toc: yes
        toc_depth: 2
        number_sections: yes
        toc_float: TRUE
editor_options: 
  chunk_output_type: console
---

# Background

## Context
Simulation of simple reaction times from a picture-naming experiment. Fifty subjects name  75 different pictures (items) under four transparency conditions (i.e., 300 trials per subject; total N = 15,000). The grayness of the screen background is randomly sampled from a normal distribution for each trial. 

## Issues
    (1) Theory-guided contrast specification for four-level factor
    (2) Specification of crossed-random factor linear mixed models
    (3) Visualization of partial effects of covariate x factor interactions

## Variables in file
    Subj : subject id; random factor
    Item : item id; random factor
    Trnsp: transparency of item (levels: A = 100%, B = 75%, C = 50 %, D = 25%); 
           within-subject and within-item fixed factor
    scrb : screen background; within-subject and within-item covariate
    rt   : reaction time, dependent variable

# Setup

There are two commands to know:

+  `install.pacakges()` retrieves packages of additional commands from the web and _installs_ them permanently in the R library on your computer. This command needs to be carried out only once and you must be connected to the internet. 
+  `library()` fetches a package of additional commands from the R library on your computer and makes them available for the current session. These commands need to be carried out in each working session.

```{r setup, include=TRUE}
#install.packages(c("tidyverse", "GGally", "Rmisc", "lme4"))  # Comment / delete the command after first use
knitr::opts_chunk$set(echo = TRUE)

library(lme4)
library(Rmisc)
library(MASS)
suppressPackageStartupMessages(library(GGally))
suppressPackageStartupMessages(library(tidyverse))
```

# Read data from csv file

We read data and generate the profile of means for the four transparency conditions.

```{r}
dat <-  read_csv("Kliegl.csv", col_types="fffdd")
```

# Profile of transparency effects

```{r}
dat %>% group_by(Subj, Trnsp) %>%
        summarySEwithin(measurevar="rt", withinvars=c("Trnsp"), idvar="Subj") %>% 
        ggplot(aes(x=Trnsp, y=rt, group=1)) +
        geom_point() + geom_line() +
        geom_errorbar(aes(ymax = rt + ci, ymin = rt - ci), width=.05) +
        scale_x_discrete("Target transparency", labels=c("A = 0%", "B = 25%", "C = 50%", "D = 75%")) +
              scale_y_continuous("Reaction time [ms]", 
                                 breaks=seq(500, 700, 25), minor_breaks=seq(500, 700, 5)) + 
        coord_cartesian(ylim=c(500, 700)) + 
        theme_classic(base_size=14)
```

# Two types of correlations

After we aggregate our data to the subject level, we need to distinguish between two types of correlations (at least ...). The first type is the usual one, that is the correlation between the reaction times of the four transparency conditions. They are expected to be very large: Subjects' response times usually generalize across experimental conditions.

The second type of correlation is the correlation between Grand Mean and experimental effects and the correlations among experimental effects. For example, with four conditions, we can look at the three correlations between the differences computed from neighboring levels of the transparency factor (i.e., d1 = B-A; d2 = C-B; d3= D-C, see profile of means in above Figure). The differences are experimental effects and the differences _between_ subjects are interindividual differences in experimental effects.

In the following chunks, we compute these two sets of correlations. **An understanding of linear mixed models will be greatly facilitated if this distinction is internalized as much as possible (and also generalized to items or other random factors.)**

## From long to wide format

This R chunk illustrates how we can convert the data frame from the long 'subject x item' format (15,000 rows = 50 subjects x 4 conditions x 75 measures) to the wide 'subject' format (50 rows with 8 means). There are two steps and we use the tidyverse dialect of R. 

First,  we  `group_by()`, and `summarize()` the items for each subject within the four conditions. This yields means in 200 rows (i.e., 50 subjects x 4 conditions) and 2 columns (i.e. `rt` and `scrb`), that is a total of 400 means. (We don't need the `scrb`column here, but leave it in as an example for future use.)

Second, from here we move to the wide format (i.e., 8 measures in 50 rows). This is accomplished with a sequence of `gather()`, `unite()`, and `spread()` commands. There are ongoing activities to simplify this task; interested participants may want to google: `tidyr::pivot_wider()`  ... and let me know if I should revise this R chunk. 

In a third step, we select the four reaction-time means and the `Subj` factor, compute the Grand Mean and the three difference scores for each subject, and add these new variables to the data frame. 
 

```{r}
dat_subj <- dat %>% 
            group_by(Subj, Trnsp) %>% 
            summarize(rt = mean(rt), scrb = mean(scrb)) %>% 
            gather(key="Var", value="value", -Subj, -Trnsp) %>% 
            unite("key", 2:3) %>% 
            spread(key="key", value="value") %>% 
            ungroup %>% 
            select(ends_with("_rt")) %>% 
            mutate(GM = rowMeans(.),
                   d1_rt = B_rt - A_rt,
                   d2_rt = C_rt - B_rt,
                   d3_rt = D_rt - C_rt)
dat_subj
```


## Correlation

### Zero-order correlations

These are the correlations between the four measures. As expected, they are very high. 

```{r}
ggscatmat(dat_subj, columns = 1:4, alpha = 1/3) + 
          xlab("") + ylab("") +
          theme_bw(base_size=14) 
```


### Effect correlations

These are the correlations between Grand Mean and experimental effects and the correlations among experimental effects. In linear mixed models we estimate correlation parameters. They have the same conceptual meaning, but are corrected for unreliability of the measures. Therefore, the correlation parameters may be larger in magnitude than these correlations. 

```{r}
ggscatmat(dat_subj, columns = 5:8, alpha = 1/3) + 
          xlab("") + ylab("") +
          theme_bw(base_size=14) 
```


## Repeated-measure multiple regression (rmMRA)

There are rmMRA shortcuts that yield the `dat_subj` used above with a single or a few lines of R code. We will use them in the context of linear mixed modeling. We start with the raw data in `dat`, carry out multiple regressions of `rt` on `Trnsp` for each subject, and compute the average regression coefficients. The derived variables `GM`, `d1`, `d2`, and `d3` are a consequence of specifying a repeated contrast (sliding-difference contrast) for the `Trnsp` factor. 


```{r}
# From above
cor(dat_subj[, 5:8])

# set contrast
contrasts(dat$Trnsp) <- MASS::contr.sdif(4)

# lme4::lmList() -- traditional R style :)
cor(coef(suppressWarnings(lmList(rt ~ 1 + Trnsp | Subj, data=dat))))

# purr::map()
dat %>% split(.$Subj)  %>% 
        map(~ lm(rt ~ 1 + Trnsp, data=.)) %>% 
        map_dfr(~ as.data.frame(t(as.matrix(coef(.))))) %>% 
        cor()
```


# Exercise

## Item analysis for current data

Carry out the analogous analysis for the random factor `Item`. Do the results differ from the ones reported here for subjects? 

## Adapt for your own data

Carry out an analogous analyses for your own data. The exercise will probably be informative if you have a large number of subjects.

# Appendix

```{r}
sessionInfo()
```


