---
title: "Change attitudes and organizational innovation and happiness in HR departments"
author: "Hamid Keshavarz"
date: "2022-03-07 (updated: `r Sys.Date()`)"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: no
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

This study tried to examine the effects of change attitudes of HR department (University of Tehran) on organizational innovation and happiness. The total population of the present study was 348 employees (HR departments) of the University of Tehran. 219 (62.9%) of the respondents were male and 129 (37.1%) were female. We also know about their experience in number of years.

The questionnaire items use a 5-level Likert scale from 1 to 5 (strongly disagree = 1, disagree = 2, have no opinion = 3, agree = 4 and strongly agree = 5).

**Independent variables:** Change attitudes of HR department (AOCQ)

+ cognitive attitude (q1 to q6)
+ affective attitude (q7 to q12)
+ behavioral attitude (q13 to q18)

** Dependent variables**

+ organizational innovation (q19 to q26; Counter Innovative Behavior, 1998)
+ organizational staff happiness  (q27 to q49; Kjerulf, 2007)
    + learning (q27 to q30)
    + self-openness (q31 to q35)
    + participation (q36 to q38)
    + positive thinking (q39 to q42)
    + meaningful work (q43 to q45)
    + interesting work (q46 to q49)

# Setup

```{r}
library(tidyverse)
library(summarytools) # dfSummary(), descr(), freq()
```

# Read the data

```{r}
data <- read_csv("data/Keshavarz_1_HR.csv")
spec(data)
#view(dfSummary(data))
descr(data, stats="common")  
glimpse(data)
```

# Preprocessing

```{r}

```




# Appendix

This command ensures that somebody at some later time can reproduce your analysis.

```{r}
sessionInfo()
```